Predicting _C. difficile_ Infection Severity from the Taxonomic Composition of the Gut Microbiome

### 2023-06-01 paper/paper.qmd
Report NNB

---
title: "Predicting _C. difficile_ Infection Severity from the Taxonomic Composition of the Gut Microbiome"
bibliography: references.bib
fontsize: 11pt
date: last-modified
date-format: medium
format:
  manuscript-pdf:
    mainfont: Helvetica
    keep-tex: true
    linenumbers: true
    doublespacing: true
    lof: false
  html:
    embed-resources: true
  gfm:
    prefer-html: true
runningtitle: Predicting CDI Severity from OTUs
runningauthor: Sovacool
papertype: Preprint
author:
  - name: Kelly L. Sovacool
    orcid: 0000-0003-3283-829X
    affiliations:
      - ref: bioinf
  - name: Sarah E. Tomkovich
    affiliations:
      - ref: micro
  - name: Megan L. Coden
    affiliations:
      - ref: mcdb
  - name: Vincent B. Young
    affiliations:
      - ref: micro
      - ref: intmed
  - name: Krishna Rao
    affiliations:
      - ref: intmed
  - name: Patrick D. Schloss
    orcid: 0000-0002-6935-4275
    email: pschloss@umich.edu
    corresponding: true
    affiliations:
      - ref: micro
      - ref: ccmb
affiliations:
  - id: bioinf
    name: Department of Computational Medicine & Bioinformatics, University of Michigan
  - id: micro
    name: Department of Microbiology & Immunology, University of Michigan
  - id: mcdb
    name: Department of Molecular, Cellular, and Developmental Biology, University of Michigan
  - id: intmed
    name: Division of Infectious Diseases, Department of Internal Medicine, University of Michigan
  - id: ccmb
    name: Center for Computational Medicine and Bioinformatics, University of Michigan
#abstract: TODO
#importance: TODO
keywords: _C. difficile_ infection, supervised machine learning, gut microbiome, amplicion sequencing
---
```{r setup, eval=TRUE, echo=FALSE, cache=FALSE}
schtools::set_knitr_opts()
knitr::opts_chunk$set(dpi = 600)
```
```{r load_data, message = FALSE}
library(here)
library(kableExtra)
library(knitr)
library(tidyverse)

@@

Current clinical guidelines specify vancomycin and fidaxomicin as the standard
antibiotics to treat CDI, with a preference for fidaxomicin due to its higher
rate of sustained resolution of CDI and lower rate of recurrence
[@stuart_clinical_2021].
The NNTs of fidaxomicin for sustained resolution and prevention of recurrence are
each estimated to be 10 [@long_oral_2022;@tashiro_oral_2022].
However, fidaxomicin is considerably more expensive than vancomycin.
If fidaxomicin were shown to reduce the risk of severe CDI outcomes, it could be
preferentially prescribed to patients predicted to be at risk, while prescribing
vancomycin to low-risk patients.
If we assume that the superior efficacy of fidaxomicin for sustained resolution
and reduced recurrence also translates to reducing the risk of severe outcomes,
we can pair the NNT of fidaxomicin with the NNS of OTU-based prediction models
to estimate the NNB.

We determined the 95th percentile of risk for each prediction model and computed
the confusion matrix and several performance metrics including the NNS at this
threshold (@tbl-risk).
Among the models predicting severe outcomes, those trained on the full
datasets performed best with an NNS of `r allcause_nns` for the all-cause
definition, `r attrib_nns` for the attributable definition, and 
`r pragmatic_nns` for the pragmatic definition.
For context, prior studies predicted CDI-attributable severity using whole
Electronic Health Record data and from a smaller set of clinician-curated
variables, achieving precision values of 0.417 (NNS = `r 1/0.417`) for the EHR
model and 0.167 (NNS = `r 1/0.167`) for the curated model [@li_using_2019;
@rao_clostridium_2015].
While the attributable definition had a worse NNS for our OTU-based models, it
did not perform worse than the prior curated model, and it is the most
clinically relevant as physician chart review increases confidence that
positively-labelled severe outcomes are due to the CDI rather than other causes.

TODO NNB
<!--
rough estimate of costs.
current: everyone gets vancomycin.
potential: patients flagged as severe get fidaxomicin. based on NNB, estimate
how much money saved in averting severe outcomes.
-->

@@

Models to guide treatment options. In the case of low-risk and non-invasive treatments
such as choosing between oral antibiotics, a higher number of false positives may
be tolerable as long as treatment cost is not unbearably high. However, for highly
invasive and irreversibly treatments such as colectomy, false positives must be minimized.
Cite studies saying fidaxomicin is cost-effective relative to vancomycin -
mentioned by @stuart_clinical_2021, e.g. @jiang_budget_2022.

@@

>>>>

---
title: "Predicting _C. difficile_ Infection Severity from the Taxonomic Composition of the Gut Microbiome"
bibliography: references.bib
fontsize: 11pt
date: last-modified
date-format: medium
prefer-html: true
format:
  manuscript-pdf:
    mainfont: Helvetica
    keep-tex: true
    linenumbers: true
    doublespacing: true
    lof: false
  html:
    embed-resources: true
runningtitle: Predicting CDI Severity from OTUs
runningauthor: Sovacool
papertype: Preprint
author:
  - name: Kelly L. Sovacool
    orcid: 0000-0003-3283-829X
    affiliations:
      - ref: bioinf
  - name: Sarah E. Tomkovich
    affiliations:
      - ref: micro
  - name: Megan L. Coden
    affiliations:
      - ref: mcdb
  - name: Vincent B. Young
    affiliations:
      - ref: micro
      - ref: intmed
  - name: Krishna Rao
    affiliations:
      - ref: intmed
  - name: Patrick D. Schloss
    orcid: 0000-0002-6935-4275
    email: pschloss@umich.edu
    corresponding: true
    affiliations:
      - ref: micro
      - ref: ccmb
affiliations:
  - id: bioinf
    name: Department of Computational Medicine & Bioinformatics, University of Michigan
  - id: micro
    name: Department of Microbiology & Immunology, University of Michigan
  - id: mcdb
    name: Department of Molecular, Cellular, and Developmental Biology, University of Michigan
  - id: intmed
    name: Division of Infectious Diseases, Department of Internal Medicine, University of Michigan
  - id: ccmb
    name: Center for Computational Medicine and Bioinformatics, University of Michigan
#abstract: TODO
#importance: TODO
keywords: _C. difficile_ infection, supervised machine learning, gut microbiome, amplicion sequencing
---
```{r setup, eval=TRUE, echo=FALSE, cache=FALSE}
schtools::set_knitr_opts()
knitr::opts_chunk$set(dpi = 600)
```
```{r load_data, message = FALSE}
library(here)
library(kableExtra)
library(knitr)
library(tidyverse)

@@

Current clinical guidelines specify vancomycin and fidaxomicin as the standard
antibiotics to treat CDI, with a preference for fidaxomicin due to its higher
rate of sustained resolution of CDI and lower rate of recurrence
[@johnson_clinical_2021].
The NNTs of fidaxomicin for sustained resolution and prevention of recurrence
are each estimated to be 10 [@long_oral_2022;@tashiro_oral_2022].
However, fidaxomicin is considerably more expensive than vancomycin.
If fidaxomicin were shown to reduce the risk of severe CDI outcomes, it could be
preferentially prescribed to patients predicted to be at risk, while prescribing
vancomycin to low-risk patients.
If we assume that the superior efficacy of fidaxomicin for sustained resolution
and reduced recurrence also translates to reducing the risk of severe outcomes,
we can pair the NNT of fidaxomicin with the NNS of OTU-based prediction models
to estimate the NNB.

To calculate a clinically-relevant NNS for these models, we computed the
confusion matrix at the 95th percentile of risk for each prediction model
(@tbl-risk).
Among the models predicting severe outcomes, those trained on the full
datasets performed best with an NNS of `r allcause_nns` for the all-cause
definition, `r attrib_nns` for the attributable definition, and 
`r pragmatic_nns` for the pragmatic definition.
For context, prior studies predicted CDI-attributable severity using whole
Electronic Health Record data and from a smaller set of clinician-curated
variables, achieving precision values of 0.417 (NNS = `r ehr_nns`) for the EHR
model and 0.167 (NNS = `r curated_nns`) for the curated model at the 95th
percentile of risk [@li_using_2019;@rao_clostridium_2015].
Multiplying the NNS of the OTU-based models by the estimated NNT of 
`r fdx_nnt` for fidaxomicin yields NNB values of `r allcause_fdx_nnb` 
for all-cause severity, `r attrib_fdx_nnb` for attributable severity,
and `r pragmatic_fdx_nnb` for pragmatic severity.
Thus, in a hypothetical scenario where these assumptions about fidaxomicin were
correct, between `r min_fdx_nnb` and `r max_fdx_nnb` patients would need to be
predicted to experience a severe outcome and be treated with fidaxomicin in
order for one patient to benefit.
As the NNS values were computed at the 95th percentile of risk (where 5% of
patients screened are predicted to experience severity), these NNB values mean
that `r min_screen` to `r max_screen` total CDI patients would need to be
screened by an OTU-based prediction model in order for one patient to benefit by
treatment guided by the model.
For comparison, the NNB for pairing the prior EHR-based model with fidaxomicin
would be `r ehr_fdx_nnb` with `r ehr_screen` total CDI patients screened for
one patient to benefit.
These estimates represent a proof-of-concept demonstration of the potential
value of deploying severity prediction models to guide clinicians' treatment
decisions.

<!--
rough estimate of costs.
current: everyone gets vancomycin.
potential: patients flagged as severe get fidaxomicin. based on NNB, estimate
how much money saved in averting severe outcomes.
-->

@@

Models to guide treatment options. In the case of low-risk and non-invasive treatments
such as choosing between oral antibiotics, a higher number of false positives may
be tolerable as long as treatment cost is not unbearably high. However, for highly
invasive and irreversibly treatments such as colectomy, false positives must be minimized.
Cite studies saying fidaxomicin is cost-effective relative to vancomycin -
mentioned by @johnson_clinical_2021, e.g. @jiang_budget_2022.

@@

Models predicting the pragmatic definition yielded the best NNS.
While the attributable definition had a worse NNS for our OTU-based models, it
did not perform worse than the prior curated model, and it is the most
clinically relevant as physician chart review increases confidence that
positively-labelled severe outcomes are due to the CDI rather than other causes.
